---
title: "Predicting Genetic Disorders"
author: "Bridget Manu"
date: "2024-02-29"
output: 
   pdf_document:
     latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```


```{r setup, include=FALSE}

library(ggplot2)

```

Data was collected from 27 different hospitals
There are 3 different genetic disorders
There are 9 sub genetic disorder classes

```{r}
data1 <- read.csv("C:\\Users\\Bridget Manu\\OneDrive\\Desktop\\Math\\Stats\\Project\\train_genetic_disorders.csv")

head(data1)
unique(data1$Patient.Age)
unique(data1$Institute.Name)
unique(data1$Genetic.Disorder)
unique(data1$Disorder.Subclass)


```

The age of the patients ranges from 0 to 14
```{r}
# visualize ages of patients

ggplot(data = data1, mapping = aes(x = Patient.Age)) + 
  geom_pie( na.rm = TRUE, bins = 35)

# visualizing frequency of each genetic disorder in the dataset

ggplot(datasub, aes(x="", y = Genetic.Disorder, fill = Genetic.Disorder,
                    color = Genetic.Disorder)) +
  geom_bar(width = 1, stat = "identity") +
  coord_polar("y", start=0) +
  theme_void() +
  labs(title = "Pie Chart of the Proportion of the Genetic Disorders")

ggplot(data = datas, mapping = aes(x = Genetic.Disorder,
                                   color = Disorder.Subclass,
                                   fill = Disorder.Subclass),  na.rm = TRUE) + 
  geom_bar(position = "dodge") + 
  facet_wrap(~ Genetic.Disorder, nrow = 3) 

ggplot(data = datas, mapping = aes(x = Genetic.Disorder, color = Gender,
                                   fill = Gender)) + 
  geom_bar(na.rm = TRUE, position = "dodge") +
  labs(title = "Frequency of each Genetic disorder among the Genders",
       x = "Genetic Disorders", y = "Frequency")


ggplot(data = datasub, mapping = aes(x = Genes.in.mother.s.side,
                                     color = Genetic.Disorder,
                                     fill = Genetic.Disorder)) + 
  geom_bar(na.rm = TRUE, position = "dodge") +
  labs(title = "Bar Graph of Frequency of each Disorders vs. Presence",
       x = "Presence of Genes in Mother's side", y = "Frequency")
ggplot(data = datas, mapping = aes(x = Inherited.from.father,
                                   color = Genetic.Disorder,
                                   fill = Genetic.Disorder)) + 
  geom_bar(na.rm = TRUE, position = "dodge") +
  labs(title = "Bar Graph of Frequency of each Disorders vs. Presence",
       x = "Inherited from Father", y = "Frequency")




```


```{r}
# creating a subset of main dataframe

datasub <- subset(data1, data1$Genetic.Disorder == "Mitochondrial genetic inheritance disorders" | data1$Genetic.Disorder == "Multifactorial genetic inheritance disorders" | data1$Genetic.Disorder == "Single-gene inheritance diseases")
                        
head(datasub)

```

Data Cleaning for the Decision tree

MFGID - Multifactorial genetic inheritance disorders
SGID - Single-gene inheritance diseases
MGID - Mitochondrial genetic inheritance disorders

 "Disorder.Subclass", "Genes.in.mother.s.side", "Maternal.gene", "Paternal.gene", "Inherited.from.father"
```{r}
not_needed <- c("Patient.Id", "Patient.First.Name", "Patient.Age",
                "Patient.First.Name", "Family.Name", "Father.s.name",
                "Institute.Name", "Location.of.Institute", "Parental.consent", "Place.of.birth", "Status", "Blood.cell.count..mcL.", "Heart.Rate..rates.min", "Follow.up", "Assisted.conception.IVF.ART", "White.Blood.cell.count..thousand.per.microliter.", "Mother.s.age", "Father.s.age", "No..of.previous.abortion", "Respiratory.Rate..breaths.min.", "Test.1", "Test.2", "Test.3", "Test.4", "Test.5", "Gender", "Birth.asphyxia", "Autopsy.shows.birth.defect..if.applicable.", "Folic.acid.details..peri.conceptional.", "H.O.serious.maternal.illness", "H.O.radiation.exposure..x.ray.", "H.O.substance.abuse",
"History.of.anomalies.in.previous.pregnancies", "Blood.test.result", "Birth.defects")

datasub <- datasub[, !(names(datasub) %in% not_needed)]

# simplify the genetic disorders by their abbreviations

datasub$Genetic.Disorder[which(datasub$Genetic.Disorder == "Multifactorial genetic inheritance disorders")] <- "MFGID"

datasub$Genetic.Disorder[which(datasub$Genetic.Disorder == "Single-gene inheritance diseases")] <- "SGID"

datasub$Genetic.Disorder[which(datasub$Genetic.Disorder == "Mitochondrial genetic inheritance disorders")] <- "MGID"

datasub$Disorder.Subclass[datasub$Disorder.Subclass == ""] <- NA

# factor the symptom variables into categorical variables

datasub$Symptom.1 <-factor(as.character(datasub$Symptom.1), levels = c("0", "1"), labels = c("Not present", "Present"))
datasub$Symptom.2 <-factor(as.character(datasub$Symptom.2), levels = c("0", "1"), labels = c("Not present", "Present"))
datasub$Symptom.3 <-factor(as.character(datasub$Symptom.3), levels = c("0", "1"), labels = c("Not present", "Present"))
datasub$Symptom.4 <-factor(as.character(datasub$Symptom.4), levels = c("0", "1"), labels = c("Not present", "Present"))
datasub$Symptom.5 <-factor(as.character(datasub$Symptom.5), levels = c("0", "1"), labels = c("Not present", "Present"))

# convert all char variables to factor

datasub[] <- lapply(datasub, function(x) {
  if(is.character(x)) return(as.factor(x))
  return(x)
})

# replaces all empty spaces with NA
datasub[] <- lapply(datasub, function(x) {
  if(is.factor(x)) x[x == ""] <- NA
  return(x)
})

# replaces all "-" with NA
datasub[] <- lapply(datasub, function(x) {
  if(is.factor(x)) x[x == "-"] <- NA
  return(x)
})

# replaces all Non applicable with NA
datasub[] <- lapply(datasub, function(x) {
  if(is.factor(x)) x[x == "Not applicable"] <- NA
  return(x)
})

# replaces all Not available with NA
datasub[] <- lapply(datasub, function(x) {
  if(is.factor(x)) x[x == "Not available"] <- NA
  return(x)
})

# replaces all No record with NA
datasub[] <- lapply(datasub, function(x) {
  if(is.factor(x)) x[x == "No record"] <- NA
  return(x)
})

datasub <- na.omit(datasub)

head(datasub)

```

Split test data and train data

```{r}
set.seed(1)

# Use 80% of the dataset as the training set and 20% as the test set

sample <- sample(c(TRUE, FALSE), nrow(datasub), replace=TRUE, prob=c(0.8,0.2))
train <- datasub[sample, ]
testing <- datasub[!sample, ]


```

```{r}

install.packages("rpart.plot")
library(rpart.plot)
library(rpart)

# Decision tree for the trained data

tree <- rpart(Genetic.Disorder ~ Genes.in.mother.s.side + Inherited.from.father + Maternal.gene + Paternal.gene + Symptom.1 + Symptom.2 + Symptom.3 + Symptom.4 + Symptom.5, data = train, method = "class")

# Plot the decision tree

prp(tree, extra = 0, nn = TRUE, fallen.leaves = FALSE, varlen = 0, faclen = 0)



```
```{r}
# Preparing the test data for prediction

not <- c("Genetic.Disorder", "Disorder.Subclass")

train1 <- train[, !(names(train) %in% not)]
testing1 <- testing[, !(names(testing) %in% not)]

```



Test case data set
```{r}
# Load the caret package
library(caret)


# predicts genetic disorder of test_case data using the trained decision tree
prediction <- predict(tret, testing1, type = "class")
prediction <- predict(tret, train1, type = "class")

# creates the confusion matrix
cm1 <- confusionMatrix(prediction, train$Genetic.Disorder)
cm1




```

```{r}
test <- rpart(Genetic.Disorder ~ Genes.in.mother.s.side + Inherited.from.father + Maternal.gene + Paternal.gene + Symptom.1 + Symptom.2 + Symptom.3 + Symptom.4 + Symptom.5,
              data = testing, method = "class")
prp(test)
prp(test, extra = 0, nn = TRUE, fallen.leaves = FALSE, varlen = 0, faclen = 0)


```


```{r}
tret <- rpart(Genetic.Disorder ~ Genes.in.mother.s.side + Inherited.from.father + Maternal.gene + Paternal.gene + Symptom.1 + Symptom.2 + Symptom.3 + Symptom.4 + Symptom.5, data = datasub, method = "class",
                       control = rpart.control(minsplit = 30))

rpart.plot(tret)

tres <- rpart(Genetic.Disorder ~ Genes.in.mother.s.side + Inherited.from.father + Maternal.gene + Paternal.gene + Symptom.1 + Symptom.2 + Symptom.3 + Symptom.4 + Symptom.5, data = testing, method = "class",
                       control = rpart.control(minsplit = 30))

rpart.plot(tres)

```



